A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2023-12-22 paper/paper.tex
Set up ott to type set set theory

\section{Semantic Typing}
\label{sec:logreldep}
\begin{figure}[h]
  \begin{equation*}
    \begin{split}
      [-]_= &:: \mathit{Term} \rightarrow \mathit{Nat} \rightarrow \mathcal{P}(\mathit{Term})\\
      [ [[Void]] ]_i &= \emptyset \\
      [ [[Bool]] ]_i &= \{ [[a]] \mid [[a =>+ true]] \text{ or } [[a =>+ false]] \} \\
      [ [[a ~ b : A]] ]_i &= \{ [[p]] \mid [[p =>+ refl]] \text{ and } [[a<=>b]] \}\\
      [ [[Pi A B]] ]_i &= \{ [[b]] \mid \forall [[a]], \text{ if } [[a]] \in [ [[A]] ]_i\text{, then } [[b a]] \in [ [[B {a}]] ]_i \} \\
      [ [[Set j]] ]_i &= \{ [[A]] \mid [[j < i]] \text{ and } [[A]] \in dom([-]_j) \}\\
      [ [[ A ]] ]_i &= \{ [[a]] \mid \exists [[B]], [[a]] \in[ [[B]] ]_i \text{ and } [[A => B]] \}
    \end{split}
  \end{equation*}
  \caption{Logical relation for \lang}
  \label{fig:logrel}
\end{figure}

In this section, we show the definition of the logical relation as a
partial function in Figure~\ref{fig:logrel}. The logical relation can
be viewed as a partial function that takes a closed type,
a natural number $[[i]]$ representing the universe level, and
returns a set of closed terms that \emph{semantically} inhabit the
closed type, though we do not strictly enforce the closedness of
either the type or the term.

The claim that $[ - ]_=$ is a partial function is not immediately
obvious due to the last equation in its definition. The last equation
ensures that the if some type $[[A]]$ is in the domain of the
interpretation function, then any type $[[B]]$ that steps into $[[A]]$
has the same interpretation as $[[A]]$. For example,
since $[[if true Bool Void]]$ steps into $[[Bool]]$, we can say that
$[[if true Bool Void]]$ is semantically inhabited by terms that
reduces down to either $[[true]]$ or $[[false]]$ through the
$[[Bool]]$ case of the interpretation function.

However, due to the non-deterministic nature of parallel reduction,
given some terms $[[A]]$, $[[B0]]$, $[[B1]]$ such that $[[A => B0]]$
and $[[A => B1]]$, the meaning of $[[A]]$ could either be $[ [[B0]] ]_i$
or $[ [[B1]] ]_i$. In fact, the definedness of $[ [[B0]] ]_i$ and $[
[[B1]] ]_i$
should not be taken as granted either. It is unclear for now whether
the definedness of one implies the other.

Fortunately,

>>>>

\section{Semantic Typing}
\label{sec:logreldep}
\begin{figure}[h]
  \begin{equation*}
    \begin{split}
      [-]_= &:: \mathit{Term} \rightarrow \mathit{Nat} \rightarrow \mathcal{P}(\mathit{Term})\\
      [ [[Void]] ]_i &= \emptyset \\
      [ [[Bool]] ]_i &= \{ [[a]] \mid [[a =>+ true]] \text{ or } [[a =>+ false]] \} \\
      [ [[a ~ b : A]] ]_i &= \{ [[p]] \mid [[p =>+ refl]] \text{ and } [[a<=>b]] \}\\
      [ [[Pi A B]] ]_i &= \{ [[b]] \mid \forall [[a]], \text{ if } [[a]] \in [ [[A]] ]_i\text{, then } [[b a]] \in [ [[B {a}]] ]_i \} \\
      [ [[Set j]] ]_i &= \{ [[A]] \mid [[j < i]] \text{ and } [[A]] \in dom([-]_j) \}\\
      [ [[ A ]] ]_i &= \{ [[a]] \mid \exists [[B]], [[a]] \in[ [[B]] ]_i \text{ and } [[A => B]] \}
    \end{split}
  \end{equation*}
  \caption{Specification of the logical relation for \lang}
  \label{fig:logrel}
\end{figure}

In this section, we present the definition of the logical relation as a
partial function.
Figure~\ref{fig:logrel} shows the specification
of the logical relation that we are about to define. The specification
itself does not immediately induce a definition as it is unclear
whether there exists a smallest partial function that satisfies the
specification.

The construction of a function that meets the specification in
Figure~\ref{fig:logrel} can be broken down into two steps. To build an
infinite hierarchy of universes, we first define a recursor that constructs
the interpretation of the universe level $[[n]]$ given a parameterized
interpretation function for universe levels that are strictly less
than $[[n]]$.
Next, we use the well-foundedness of the less than relation to tie the
knot and define the full interpretation function that meets the
specification.
The resulting definition is then shown to be functional and thus
justifies the notation $[ [[A]] ]_i$ since we know there would be an
unambiguous output corresponding to the type $[[A]]$, assuming that
$[[A]]$ is in the domain of the partial function.
